The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-05-21 dictionary.tex
Removing speculation in spectral sequences section, other fixes

\title[Lost in Translation: The function--sheaf dictionary]{Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields}

@@

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes over finite fields.
  To do so, we introduce the monoidal category of quasicharacter sheaves on
  smooth commutative group schemes $G$ over $\mathbb{F}_q$
  and show that the Trace of Frobenius is a functorial surjection
  from the group of isomorphism classes of quasicharacter sheaves on $G$
  onto the character group of $G(\mathbb{F}_q)$.
  If $G$ is connected, the kernel of this surjection is trivial.
  In $G$ is etale, however, the kernel is generally non-trivial,
  leading to the interesting notion of invisible quasicharacter sheaves.
  In this paper we find a cohomological classification of invisible quasicharacter sheaves on all smooth commutative group schemes over $\mathbb{F}_q$,
  and also explain how to make invisible quasicharacter sheaves disappear,
  should the notion prove too troubling.
   \end{abstract}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
This paper gives a cohomological description of invisible quasicharacter sheaves on $G$
and a trick to make invisible quasicharacter sheaves disappear,
should the notion prove too troubling.

\subsection*{Main result}

In order to describe the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there \emph{exists} an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask how the function--sheaf dictionary generalizes to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$ over a finite field $\Fq$.
To do this we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and also keep track of the isomorphisms \eqref{introbox}, leading to the notion of {\it quasicharacter sheaves on $G$}.

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.

The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
If $G$ is a connected commutative algebraic group then
all quasicharacter sheaves on $G$ are visible.
In this paper we find a simple cohomological description
of invisible quasicharacter sheaves on any smooth commutative
group scheme $G$ over $\Fq$.
In any case, the canonical section of the Trace of Frobenius allows
one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$;
consequently, quasicharacter sheaves geometrize and categorify
characters of $G(\Fq)$.

\subsection*{Illustrative examples}

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to smooth commutative group schemes over $\Fq$.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
In particular, we will see how to uniquely geometrize the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$, which has unbounded image,
using quasicharacter sheaves on the etale group scheme $\ZZ$ with trivial Galois action.
All quasicharacter sheaves on this group are visible.
Likewise, the dictionary is perfect for the
etale group scheme $\ZZ$ with Galois action determined by $\Frob{}(n) = -n$.
By contract, the etale group scheme $\ZZ \times \ZZ$ with trivial Galois action
 does admit invisible quasicharacter sheaves -- there are non-trivial quasicharacter sheaves on this group
with trivial trace of Frobenius.\todo{Under construction}
(Move this example: Define $g: \ZZ\times \ZZ \to \EEx$ by $g((n_1,m_1), (n_2, m_2)) = \ell^{m_1-m_2 + n_1-n_2}$.
Then use the tehniques of Section~\ref{sec:etale} to build a non-trivial character sheaf from the pair $(f=1,g)$ with trivial trace of Frobenius.)

\subsection*{Motivation and Application}

Our interest in the function--sheaf dictionary for smooth commutative group schemes over $\Fq$, especially etale commutative group schemes, comes from an application to $p$-adic representation theory.
In \cite{cunningham-roe:13a} we consider quasicharacter sheaves on the
Greenberg transform of the N\'eron model of an algebraic torus over an arbitrary local field.
The ability to work with non-connected group schemes,
of which the component group of the special fibre of the Neron model of a torus is an important example,
plays a crucial role in this application,
since these group schemes are not connected in general.
This process of creating a category from the group of quasicharacters of such a torus
informs our choice of the term \emph{quasicharacter sheaf} here.

\subsection*{Comparison with other character sheaves}

We also want to situate our terminology in a historical context.
It is not uncommon to refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}.
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
We note that there are no invisble quasicharacter sheaves on such groups.

\subsection*{Structure of the paper}

%We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves equipped with an isomorphism of the form  \eqref{introbox}.
Section~\ref{sec:Frob} introduces Trace of Frobenius,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on Trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{sec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves, in this section
we also give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$, where $\Weil{}$ is the Weil group for $\Fq$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
is a functorial isomorphism for $\QC$-schemes.

@@

\subsection{Notations}\label{sec:QCschemes}

@@

\subsection{Main definition}\label{sec:category}

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(QC.1)}{QC.1} $\gqcs{L}$ is a rank~one $\ell$-adic local system on $\bG$, by which we mean
a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected
component of $\bG$, whose stalks are one-dimensional $\EE$-vector spaces;
\labitem{(QC.2)}{QC.2} $\mu: \bm^* \gqcs{L} \to \LxL$ is an isomorphism of
sheaves on $\bGxG$ such that the following diagram commutes,
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
\labitem{(QC.3)}{QC.3} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with
  $\mu$ in the sense that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\section{Two special subcategories of quasicharacter sheaves}%\label{sec:bounded}

@@

\section{Spectral sequences}\label{sec:spectral}

%\subsection{Invisible quasicharacter sheaves}\label{sec:invisible}

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not a connected algebraic group.
Our main tool for understanding this phenomenon
is a group homomorphism from $\QCiso{G}$ to $\Hh^2(E^\bullet_G)$, defined in this section. In Section~\ref{sec:restriction} we will see that this homomorphism is an isomorphism for all smooth commutative group schemes over $\Fq$.

[Things I need from this section:
definition of $E^\bullet_G$ (Section~\ref{sec:E});
the short exact sequence attached to $\Hh^2(E^\bullet_G)$ (Section~\ref{sec:SES});
the exactness of all sequences in the diagram below (Section~\ref{sec:front});
the map $S_G$ from $\QCiso{G}$ to $\Hh^2(E^\bullet_G)$ (Section~\ref{sec:S} done) commutes with everything in sight;
that $\Hh^0(\Weil,\Hh^2({\bar G},\EEx))=0$ if $G$ is a connected algebraic group over $\Fq$ (Section~\ref{sec:connected-iso}).]

@@

Define $\Hh^2(E^\bullet_G)$.

Do we have
 \[
  \begin{tikzcd}[column sep=30]
    ? \cdots \arrow{r}
%    & \Hh^1(E^\bullet_{A}) \arrow{r}
    & \Hh^2(E^\bullet_{C}) \arrow{r}
    & \Hh^2(E^\bullet_{B}) \arrow{r}
    & \Hh^2(E^\bullet_{A}) \arrow{r}
%    & \Hh^3(E^\bullet_{C}) \arrow{r}
    & \cdots ? \\
  \end{tikzcd}
 \]

[Question: If $C$ is etale, then $\Hh^3(E^\bullet_{C}) =0?$]

\[\Hh^*(E^\bullet_{G}) = \Hh^*({\bar G}\rtimes \Weil{},\EEx)\]

@@

 \[
  \begin{tikzcd}
    0 \arrow{r}
    & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r}
    & \Hh^2(E^\bullet_G) \arrow{r}
    & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r}
    & 0
  \end{tikzcd}
 \]

\subsection{Splitting?}\label{sec:split}

 \[
  \begin{tikzcd}
    0 \arrow{r}
    & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r}
    & \Hh^2(E^\bullet_G) \arrow{r}
    \arrow[bend right,dotted]{l}
    & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r}
    \arrow[bend right,dotted]{l}
    & 0
  \end{tikzcd}
 \]

\subsection{Front}\label{sec:front}

 \[
  \begin{tikzcd}[row sep=20, column sep=15]
    &  ?  & 0 \arrow{d} & 0\arrow{d} & \\
0 \arrow{r} \arrow{d}   & \Hh^0(\Weil{},\Hh^2(C,\EEx)) \arrow{r} \arrow{d} & \Hh^0(\Weil{},\Hh^2(B,\EEx)) \arrow{r} \arrow{d}
    & \Hh^0(\Weil{},\Hh^2(A,\EEx)) \arrow{d} \arrow{r} & ? \\
    %\arrow{rr}  &   & 0 \\
    ? \arrow{r}
    & \Hh^2(E^\bullet_{C}) \arrow{r} \arrow{d}
    & \Hh^2(E^\bullet_{B}) \arrow{r} \arrow{d}
    & \Hh^2(E^\bullet_{A}) \arrow{r} \arrow{d}
    & ? \\
    %\arrow{rr}&& 0\\
? \arrow{r}  & \Hh^1(\Weil{},\Hh^1(C,\EEx)) \arrow{d} \arrow{r}
    & \Hh^1(\Weil{},\Hh^1(B,\EEx)) \arrow{d} \arrow{r}
    & \Hh^1(\Weil{},\Hh^1(A,\EEx)) \arrow{d} \arrow{r}  &  ? \\
    &  0 & 0 & 0 & \\
  \end{tikzcd}
 \]

@@

\begin{proposition}\label{prop:ST}
 The following diagram commutes:
\[
\begin{tikzcd}
 0 \arrow{r}
 & \ker \TrFrob{G} \arrow{r} \arrow{d}
 & \QCiso{G}\arrow{d}{S_G} \arrow{r}{\TrFrob{G}}
 & G(\Fq)^* \arrow{d}{\iso} \arrow{r}
 & \coker \TrFrob{G} \\
 0 \arrow{r}
 & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r}
 & \Hh^2(E^\bullet_G) \arrow{r}
 & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx))  \arrow{r}
 & 0.
 \end{tikzcd}
 \]
\end{proposition}

\begin{proof}
We've discussed the fact that the square on the right commutes, using Lemma~\ref{lem:dual-inv}. I'll add that, later.
So, if $\qcs{L} \in \QC(G)$ and $S_G(\qcs{L}) = [\alpha\oplus\beta]$  then $\trFrob{\qcs{L}} =1$ implies $[\beta]=0$. In Section~\ref{sec:SES} we identified\todo{ ... will identify ...} $\{ [\alpha\oplus\beta] \in \Hh^2(E^\bullet_G) \tq [\beta]=0 \}$ with $\Hh^0(\Weil{},\Hh^2({\bar G},\EEx))$.
\end{proof}

%We begin with a simple result relating duals, invariants and coinvariants.

@@

\subsection{Invisible quasicharacter sheaves}

\begin{definition}\label{def:invisible}
We say $\qcs{L}\in \QC(G)$ is \emph{visible} if the projection of $S_G(\qcs{L})$ to $\Hh^0(\Weil{}, \Hh^2({\bar G},\EEx))$ is trivial; otherwise, we say $\qcs{L}$ is \emph{invisible}.
\end{definition}

This terminology will be justified later when we show that a non-trivial quasicharacter sheaf $\qcs{L}$ on $G$ is invisible if and only if $\trFrob{\qcs{L}}$ is trivial.

\subsection{Looking back to connected commutative algebraic groups}\label{sec:connected-iso}

We end this Section~\ref{sec:spectral} by appling the tools developed here to the case when $G$ is a connected commutative algebraic group over $\Fq$.

>>>>

\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields}

@@

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes over finite fields,
  in which case some interesting new features of the dictionary arise.
   \end{abstract}

@@

%\subsection*{Overview}

%Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
%The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
%The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
%
%In this paper we show how this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, in which case some interesting new features of the dictionary arise.

%\subsection{The main result}

%In order to describe the main result of this paper with more precision,
Let us briefly review how Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there \emph{exists} an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.
%
Although hidden in this discussion, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.
In particular, when the isomorphism \eqref{introbox} exists for $\qcs{L}$, it is essentially unique.

Now let $G$ be any smooth commutative group scheme over $\Fq$.
Two new features must be addressed in order to geometrize characters of $G(\Fq)$.
First, we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and, second, we must keep track of the isomorphisms \eqref{introbox}.
Both issues are addressed in the main definition of this paper:
A \emph{(tight) quasicharacter sheaf} on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
A \emph{(loose) quasicharacter sheaf} on $G$ is a Weil sheaf $(\gqcs{L},\phi)$ on $G$
for which $(\gqcs{L},\mu,\phi)$ is a tight quasicharacter sheaf on $G$, for some $\mu$.
The difference between these two notions is detected by $\Hh^2(\pi_0({\bar G}),\EEx)$, where $\pi_0({\bar G})$ is the component group for ${\bar G}$.
%namely, when $\Hh^2(\pi_0({\bar G}),\EEx)$ is non-trivial, if $(\gqcs{L},\phi)$ is a (loose) quasicharacter sheaf on $G$ then there are many $\mu$ for which $(\gqcs{L},\mu,\phi)$ is a (tight) quasicharacter sheaf on $G$;
%
 In the main result of this paper, Theorem~\ref{thm:snake},
we form a rigid monoidal category $\lQC(G)$ of (loose) quasicharacter sheaves on $G$
and show that the Trace of Frobenius provides a functorial group isomorphism
\begin{equation}
\lQCiso{G} \to \Hom(G(\Fq), \EEx).
\end{equation}

%\subsection*{Illustrative Examples}

%Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to smooth commutative group schemes over $\Fq$.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
For example, we will see how to geometrize the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$, which has unbounded image,
using quasicharacter sheaves on the etale group scheme $\ZZ$ with trivial Galois action.
%Likewise, the dictionary is perfect for the etale group scheme $\ZZ$ with Galois action determined by $\Frob{}(n) = -n$. By contract, the etale group scheme $\ZZ \times \ZZ$ with trivial Galois action does admit invisible quasicharacter sheaves -- there are non-trivial quasicharacter sheaves on this group with trivial trace of Frobenius.\todo{Under construction} (Move this example: Define $g: \ZZ\times \ZZ \to \EEx$ by $g((n_1,m_1), (n_2, m_2)) = \ell^{m_1-m_2 + n_1-n_2}$. Then use the tehniques of Section~\ref{sec:etale} to build a non-trivial character sheaf from the pair $(f=1,g)$ with trivial trace of Frobenius.)

%\subsection*{Motivation and Application}

Our interest in the function--sheaf dictionary for smooth commutative group schemes over $\Fq$, especially etale commutative group schemes, comes from an application to $p$-adic representation theory.
In \cite{cunningham-roe:13a} we consider quasicharacter sheaves on the
Greenberg transform of the N\'eron model of an algebraic torus over an arbitrary local field.
The ability to work with non-connected group schemes,
of which the component group of the special fibre of the Neron model of a torus is an important example,
plays a crucial role in this application,
since these group schemes are not connected in general.
This process of creating a category from the group of quasicharacters of such a torus
informs our choice of the term quasicharacter sheaf, here.

%\subsection*{Comparison with other character sheaves}

We also want to situate our terminology in a historical context.
It is not uncommon to refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}.
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
The interesting new features pertaining to Weil sheaves and the non-triviality of $\Hh^2(\pi_0({\bar G}),\EEx)^{\Frob{G}}$ do not arise in this case.

%\subsection*{Structure of the paper}

We close the introduction with a brief description of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves equipped with an isomorphism of the form \eqref{introbox}.
Section~\ref{sec:Frob} introduces Trace of Frobenius,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on Trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{sec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and find a cohomological interpretation
of quasicharacter sheaves in this case.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
is a functorial isomorphism from isomorphism classes of (loose) quasicharacter sheaves on $G$
to characters of $G(\Fq)$.

@@

\subsection{Notations}\label{sec:notation}

@@

\subsection{(tight) quasicharacter sheaves}\label{sec:category}

\begin{definition}\label{def:QC}
A \emph{(tight) quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(QC.1)}{QC.1} $\gqcs{L}$ is a rank~one $\ell$-adic local system on $\bG$, by which we mean
a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected
component of $\bG$, whose stalks are one-dimensional $\EE$-vector spaces;
\labitem{(QC.2)}{QC.2} $\mu: \bm^* \gqcs{L} \to \LxL$ is an isomorphism of
sheaves on $\bGxG$ such that the following diagram commutes,
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
\labitem{(QC.3)}{QC.3} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with
  $\mu$ in the sense that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\section{Two subcategories of quasicharacter sheaves}%\label{sec:bounded}

@@

\'Etale group schemes form a counterpoint to connected groups,
since the component group of a smooth commutative group scheme
is an \'etale group scheme \cite{vdGeer-Moonen:AbelianVarieties}*{III, \S 4}.
However, the function--sheaf dictionary is substantially different in the two cases.
To see why, we begin by specializing the definition of quasicharacter sheaves to the case
of etale commutative group schemes,
using the equivalence $G \mapsto G(\bFq)$
from the category of \'etale group schemes over $\Fq$ to the category of groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.
Under this equivalence, a (tight) quasicharacter sheaf on an etale commutative group scheme $G$ over $\Fq$ is given by the data of:
\begin{enumerate}
 \labitem{(qc.1)}{qc.1} an indexed set of one-dimensional
  $\EE$-vector spaces $\gqcs{L}_x$, as $x$ runs over
  $G(\bFq)$;

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of tight quasicharacter sheaves on $G$ when $G$ is not a connected algebraic group.
Our main tool for understanding this phenomenon
is a group homomorphism from $\QCiso{G}$ to $\Hh^2(E^\bullet_G)$, defined in Section~\ref{sec:S}, for which the next two sections are preparation.

@@

Define $\Hh^2(E^\bullet_G)$. \todo{David, I need some help here.}

@@

 \[
  \begin{tikzcd}
    0 \arrow{r}
    & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r}
    & \Hh^2(E^\bullet_G) \arrow{r}
    & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r}
    & 0
  \end{tikzcd}
 \]
\todo{Again, I hope you can write this section, David.}

@@

\begin{remark}
Although we will only use the group homomorphism $S_G : \QCiso{G}\to \Hh^2(E^\bullet_G)$ when $G$ is an etale commutative group scheme, the function is defined for all smooth commutative group schemes $G$.
However, the main result of the next section, Proposition~\ref{prop:etale-iso}, is false in that context.
\end{remark}

\subsection{$S$ and $T$}

In this section we find the relation between the group homomorphisms $\TrFrob{G}$ and $S_G$.
%
We begin with a simple result relating duals, invariants and coinvariants.

@@

\begin{proposition} \label{prop:etale-iso}
 If $G$ is an \'etale commutative group scheme over $\Fq$
  then the map $\QCiso{G}\to \Hh^2(E^\bullet_G)$ from Section~\ref{sec:S} induces an isomorphism of short exact sequences:
\[
\begin{tikzcd}
 0 \arrow{r} & \ker \TrFrob{G} \arrow{d} \arrow{r} & \QCiso{G}\arrow{d}{S_G} \arrow{r}{\TrFrob{G}} \arrow{r} & G(\Fq)^* \arrow{d} \arrow{r} & 0\\
  0 \arrow{r} & \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \arrow{r} & \Hh^2(E^\bullet_G) \arrow{r} & \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \arrow{r} & 0.
 \end{tikzcd}
 \]
In particular,
$\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$
is surjective.
\end{proposition}

\begin{proof}
We've discussed the fact that the square on the right commutes, using Lemma~\ref{lem:dual-inv}. \todo{David, could you add that here?}
So, if $\qcs{L} \in \QC(G)$ and $S_G(\qcs{L}) = [\alpha\oplus\beta]$  then $\trFrob{\qcs{L}} =1$ implies $[\beta]=0$. In Section~\ref{sec:SES} we identified\todo{ ... and this, too?} $\{ [\alpha\oplus\beta] \in \Hh^2(E^\bullet_G) \tq [\beta]=0 \}$ with $\Hh^0(\Weil{},\Hh^2({\bar G},\EEx))$.
